NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PE and PG lipids

### 2021-01-20 manuscriptPGPE.tex
FIx typo

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C--H bond order parameters from different MD force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs.~\ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig.~\ref{structures}B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in Fig.~\ref{structures}C). Major differences between headgroups are observed for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of PS lipids for which differences distinct distributions are observed particularly around phosphate group. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.

>>>>

To resolve the conformational ensembles of lipid headgroups, we compared the headgroup and glycerol backbone C--H bond order parameters from different MD force fields to the experimental data. None of the force fields reproduced the order parameters within experimental accuracy for any of the lipids analysed in this work (Figs.~\ref{HGorderParametersPE} and \ref{HGorderParametersPOPG}) or in our previous studies \cite{botan15,antila19}. Nevertheless, the CHARMM36 force field performs best for all the headgroups and reproduces the experimentally observed distinct order parameters of PS and PG lipids (Fig.~\ref{structures}B). To understand the structural origin of these distinct order parameters, we calculated the distributions of heavy atom dihedral angles from CHARMM36 simulations in Fig.~\ref{structures}C). Major differences between headgroups are observed for the last two dihedrals in the headgroup end,  O$_\alpha$-C$_\alpha$-C$_\beta$-N/C$_\gamma$ and P-O$_\alpha$-C$_\alpha$-C$_\beta$, which prefer \textit{gauche$^-$} conformations for PG and \textit{gauche$^+$} for PS. Rest of the dihedrals are similar between different lipids, with the exception of PS lipids for which distinct distributions are observed particularly around phosphate group. Therefore we suggest that the main differences between lipid headgroups leading to distinct order parameter occur in the choline part, while also changes in phosphate region may contribute in PS lipids, which could also explain the more rigid headgroup structures~\cite{browning80,buldt81}.
Furthermore, the angle between headgroup dipole and membrane normal decreases in the order of PG $>$ PE  $>$ PC  $>$ PS (Fig.~\ref{structures}D)). However, the differences between PC and PE in P-O$_\alpha$-C$_\alpha$-C$_\beta$ dihedral
and P--N vector dipole may be artificial as the $\beta$-carbon order parameter in PC is too negative in the CHARMM36 force field, thereby not being equal to the order parameter in PE as observed in experiments \cite{botan15}.
